---
title: Re-encoding Greek characters from 'EDH' dataset
#title: Using internal functions of 'sdam'
#date: "`r format(Sys.time(), '%d %B %Y')`"
date:  "4 March 2021" 
author: 
  - name: "Antonio Rivero Ostoic"
    affiliation: <center>CEDHAR, Aarhus University</center>
    email: <center>jaro@cas.au.dk</center>
output:
  html_document:
    theme: united
    highlight: tango
    code_folding: none
    keep_md: true
  pdf_document: default
---

<style type="text/css">
h1, h4 {
  text-align: center;
  color: DarkRed;
}
h1.title {
  font-size: 24pt;
}
h2 {
  font-size: 22pt;
}
h3 {
  font-size: 18pt;
}
p.output {
background-color: #FFFFFF;
padding: 10px;
border: 1px solid #C0C0C0;
margin-left: 0px;
border-radius: 5px;
font-family: monospace;
font-size: 11pt;
font-weight:bold;
}
</style>




```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,error=TRUE)
knitr::opts_chunk$set(comment = "")
library("sdam")
source("C:/R/AUArts/Epigraphic/src/cln4.r")
load(file="C:/R/AUArts/Rmds/data/edhdf.rda")
```

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
```



<div style="margin-bottom:60px;"> </div>


### Preliminaries
`"sdam"` is an `R` package that provides tools for performing analyses within Social Dynamics and complexity in the
Ancient Mediterranean [SDAM](https://sdam-au.github.io/sdam-au/), which is a research group based at Aarhus University. 
It allows accessing and manipulating data from the Epigraphic Database Heidelberg API, performing HTTP requests, 
visualizing interval time events, compute probabilities of existence of archaeological assemblages and artifacts, and 
produce similarity matrices by simple matching between these objects. 


<!-- ```{r sdam, echo=TRUE, eval=FALSE}
# remove old version from AU computer
remove.packages("sdam", lib="C:/Users/AUID/R/win-library/version/")
# install RC version from Github
require("devtools")
devtools::install_github("sdam-au/sdam")
```
 -->


```{r sdam0, echo=TRUE, eval=FALSE}
# install RC version from Github
devtools::install_github("sdam-au/sdam")
# or the beta version
devtools::install_github("mplex/cedhar", subdir="pkg/sdam")
```


<div style="margin-bottom:15px;"> </div>


```{r}
# load package and check version
library("sdam")
packageVersion("sdam")
```


<div style="margin-bottom:60px;"> </div>


<!-- 
```{r, datasets, echo=TRUE, eval=FALSE}
data(package="sdam")
```
```
Datasets in package `sdam´:

EDH                                                 Epigraphic Database Heidelberg Dataset
rp                                                  Roman province names and acronyms as in EDH
rpmp                                                Maps of ancient Roman provinces and Italian regions
rpmcd                                               Caption maps and affiliation dates of Roman provinces
```
<div style="margin-bottom:30px;"> </div>
For instance, the acronym names of Roman provinces found in the `EDH` dataset 
-->


<div style="margin-bottom:60px;"> </div>


## EDH people
`EDH` is a dataset that contains the texts of Latin and Latin-Greek inscriptions of the Roman Empire, which have 
been retrieved from the [Epigraphic Database Heidelberg API repository](https://edh-www.adw.uni-heidelberg.de/data/api) 
through routines `get.edh()` and `get.edhw()`. 

The following sample is from the `EDH` dataset where we extract variables with function `edhw()` 
and record this data in object `edhdf`. 

```{r, edhdf, echo=TRUE, eval=FALSE}
# people nomen tribus dates provs
edhdf <- edhw(vars=c("people","not_before","not_after","province_label"), select=c("tribus","nomen"))
```

<div style="margin-bottom:30px;"> </div>

By default, `edhw()` produces a data frame that in this case has 6 variables:

```{r, edhdfstr, size="tiny"}
# object list with its structure
ls.str(edhdf)

```

<div style="margin-bottom:60px;"> </div>



### People's *nomen* per province
`"sdam"` comes with a suite of datasets where `"rp"` containes Roman province names and acronyms as in EDH, 
and we use this data to organize the extracted information per Roman province in `edhdf`. 
In the following example, each `"rp"` name defines a component of list `edhs` by applying the `province` 
argument of function `edhw()`. 


```{r}
# load province names and acronyms as in EDH
data("rp")
# retrieve data per province
edhs <- list()
for(i in names(rp)) {
	edhs[[length(edhs)+1]] <- edhw(edhdf, province=i)
}
names(edhs) <- names(rp)
```

<div style="margin-bottom:30px;"> </div>

The length of `edhs` is the number of provinces in `rp`.

```{r}
length(edhs)
```

<div style="margin-bottom:30px;"> </div>


We check for instance the first two entries of `Achaia` whose cartographical map is produced by funtion 
`plot.map` as in the [Plotting ancient Roman provinces and regions vignette](https://sdam-au.github.io/sdam/articles/RomanProvs.html).


```{r, echo=TRUE, eval=FALSE}
# start at first province
head(edhs$Ach)
```
```{r, echo=FALSE, eval=TRUE}
utils::head(edhs$Ach)
```


```{r, echo=FALSE, eval=TRUE, out.width="35%", fig.align="center", fig.cap="Roman province of Achaia (117 AD)."}
plot.map("Ach", cap=FALSE, name=FALSE)
```



<div style="margin-bottom:60px;"> </div>


### Cleaning EDH *nomen*
Inscriptions in some provinces were written using Greek characters and, due to encoding and decoding steps in the 
retrieving of the data (perhaps only specific to MS Windows?), Greek characters are not writen properly with `get.edh()`. 
<!-- This circumstance can be avoided by activating the `clean` argument in this function that calls the internal function `cln()`. --> 

For instance, we record from `edhs` the last 50 available *nomen* entries of `Achaia` in object `Ach`, and 
a second step discards the not available information. 

```{r, echo=-4, eval=TRUE}
# last 50 available entries in Achaia
Ach <- tail(edhs$Ach$nomen, 50)
Ach <- Ach[!is.na(Ach)]
Ach
```
```{r, echo=FALSE, eval=FALSE}
stats::na.omit(utils::tail(edhs$Ach$nomen, 50))
```
`r emo::ji("confused")`

<div style="margin-bottom:30px;"> </div>


As we can see, these entries in `Ach` have a mixture of Latin extended characters and unicode code points. 
`cln()` (that stands for "cleaning") is a `sdam` internal function that allows reverting encoding 
corruption process of Greek texts that we saw in the *nomen* entries of `edhs`.


```{r, clnAch, echo=TRUE, eval=FALSE}
# clean encoding
cln(Ach)
```

Which produces these *nomen* with Greek characters:


<p class="output">
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[2]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[3]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[4]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[5]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[6]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[7]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[12]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[13]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[15]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[16]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[17]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[18]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[19]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[22]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[24])[[1]])
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[26]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[27]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(edhs$Ach$nomen, 50)[38]))
```
</p>
`r emo::ji("smile")`


&ast; By default, function `cln()` removes question marks from the input text, 
unless the `level` argument is set to `0`. It is also possible to raise 
the cleaning level of the function for other characters in the input such as `*` and `+`. 
Also note that in the `R` console output asterisks are kept when handling with 
Greek text (and the above omisions may be due to a `knitr` (or `rmarkdown`) issue. 


<!-- for(i in which(!(is.na(utils::tail(edhs$Ach$nomen, 50))))) -->
<!-- 2  3  4  5  6  7 12 13 15 16 17 18 19 22 24 26 27 38 -->

<!-- "<U+039F><U+1F50><U+03C1><U+03B9><U+03BF><U+03C2>+" -->



<div style="margin-bottom:60px;"> </div>


##### Nomen in Aegyptus

Similarly to `Ach`, entries in `Aegyptus` can be cleaned with `cln()`. 

<div style="margin-bottom:30px;"> </div>

```{r, echo=FALSE, eval=TRUE, out.width="35%", fig.align="center", fig.cap="Roman province of Aegyptus (117 AD)."}
plot.map("Aeg", cap=FALSE, name=FALSE)
```


<div style="margin-bottom:30px;"> </div>


In the following cases from Aegyptus, the entries have both Greek and Latin characters. 

```{r, echo=TRUE, eval=TRUE}
# record available nomen variable for Aegyptus
Aeg <- edhs$Aeg$nomen
Aeg <- Aeg[!is.na(Aeg)]
```
```{r, echo=TRUE, eval=TRUE}
# last five entries
tail(Aeg, 5)
```



<div style="margin-bottom:30px;"> </div>


```{r echo=TRUE, eval=FALSE}
# Greek characters in Aegyptus
cln(tail(Aeg, 5))
```


<p class="output">
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 5)[1]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 4)[1]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 3)[1]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 2)[1]))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 1)))
```
</p>


<div style="margin-bottom:60px;"> </div>


<!-- ```{r echo=TRUE, eval=FALSE}
cln(tail(Aeg, 3), level=2)
```

<p class="output">
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 3)[1], level=2))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 2)[1], level=2))
```
<br>
```{r, echo=FALSE, eval=TRUE}
knitr::asis_output(cln4(utils::tail(Aeg, 1), level=2))
```
</p> -->





<div style="margin-bottom:60px;"> </div>



### See also

* [Plotting ancient Roman provinces and regions](https://sdam-au.github.io/sdam/articles/RomanProvs.html)
<!-- * [R package `"sdam"`](https://sdam-au.github.io/sdam/) -->
* [`"sdam"` manual](https://github.com/mplex/cedhar/blob/master/typesetting/reports/sdam.pdf)
* [Social Dynamics and complexity in the Ancient Mediterranean Project](https://sdam-au.github.io/sdam-au/)



<div style="margin-bottom:60px;"> </div>




